Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-02-25 paper/sec/invariants.tex
Paper: Fix typo in Sec. 3.
minor edits (.. -> ...)

\subsection{Deriving A Simple Probabilistic Invariant}
Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,..,a_m\}$ after loop iteration $m$.

@@

Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
\]
Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.

@@

Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
This leads to a similar inequality:
\begin{align}
  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
\end{align}
with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows:
%TODO: should we use exp(...) everywhere instead of mixing with e^ ? EK: That was the previous verison, main issue is the chain looks very lopsided, with almost all of the terms being on the right edge. (It looked too ugly.)
% The convention is now, if exp is applied to a random variable we use the long notation and otherwise the power notation, which might actually be (a bit) helpful.
\begin{eqnarray*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq & % Is not equal when |A|=0.
  \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) \exp( t p^{-1} X_s) \right] \\
  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) h(p^{-1} X_s) \right] \\
  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
  & \underset{\textrm{Calculus}}{\leq} & e^{-n\varepsilon^2 / 12}
\end{eqnarray*}
where we choose $h(x) = 1+qx (e^{t/q}-1)$.
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
To use these bounds, we also have to separately estimate $\prob(p < q)$.
For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, with $q = \frac{n}{4 |A|}$.
%The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
The formalization i nthe supplementary material contains a detailed informal step-by-step proof using our approach in its appendix.
Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.

>>>>

\subsection{Deriving A Simple Probabilistic Invariant}
Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,\dots,a_m\}$ after loop iteration $m$.

@@

Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
\]
Note that we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.

@@

Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
This leads to a similar inequality:
\begin{align}
  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
\end{align}
with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows:
%TODO: should we use exp(...) everywhere instead of mixing with e^ ? EK: That was the previous verison, main issue is the chain looks very lopsided, with almost all of the terms being on the right edge. (It looked too ugly.)--- OK
% The convention is now, if exp is applied to a random variable we use the long notation and otherwise the power notation, which might actually be (a bit) helpful.
\begin{eqnarray*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq & % Is not equal when |A|=0.
  \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) \exp( t p^{-1} X_s) \right] \\
  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) h(p^{-1} X_s) \right] \\
  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
  & \underset{\textrm{Calculus}}{\leq} & e^{-n\varepsilon^2 / 12}
\end{eqnarray*}
where we choose $h(x) = 1+qx (e^{t/q}-1)$.
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
To use these bounds, we also have to separately estimate $\prob(p < q)$.
For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2023}, with $q = \frac{n}{4 |A|}$.
%The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
The formalization in the supplementary material contains a detailed informal step-by-step proof using our approach in its appendix.
Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.

### 2025-02-25 paper/sec/negdep.tex
some space saving
minor edits (.. -> ...)

\subsection{Background on Negative Dependence}
Some sets of random variables possess a property called \emph{negative association}, a generalization of independence.
The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orthant dependence.
Importantly, standard Chernoff--Hoeffding type bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
Negative association is defined as follows:
\begin{definition}
For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, i.e., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, s.t., $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
\end{definition}

\begin{definition}[Negative Association]\label{def:neg_assoc}
A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
\[
\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
\]
\end{definition}

The following proposition summarizes some of the properties for negatively associated sets of random variables.
\begin{proposition}[\cite{joagdev1983}]\label{pro:neg_dep_props}
Summary of results for negatively associated random variables.
\begin{enumerate}
\item \label{it:neg_dep_props:mult_mono} If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector valued random variables $X$ and $Y$ are independent, then the union $X_1,..,X_n,Y_1,..,Y_m$ is a set of negatively associated random variables.
\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $f_1,..,f_m : \mathbb R^n \rightarrow \mathbb R$ be all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
\item If $X_1,..,X_n$ are independent then $X_1,...,X_n$ are negatively associated.
\item A subset of a negatively associated set of random variables is again negatively associated.
\end{enumerate}
\end{proposition}

These properties illustrate the trade-off between negative association and independence.
For example, property 3 would be true for independent random variables, even without the condition of monotonicity.
%Of course, on the other hand independence is a stronger property and fewer sets of random variables are independent.
To analyze our new subsampler, the following is an important lemma about negative associated random variables.

\begin{lemma}\label{le:neg_assoc_prod}
Let $X_1,..,X_n$ be negatively associated and $f_1,..,f_n$ be all non-decreasing (or all non-increasing), non-negative functions, then
\[
  \expect \left[\prod_{i=1}^{n} f_i(X_i)\right] \leq \prod_{i=1}^{n} f_i(\expect [X_i]) \textrm{.}
\]
\end{lemma}
\begin{proof}
This follows from the definition of negative associativity (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
\end{proof}

>>>>

\subsection{Background on Negative Dependence}
Some sets of random variables possess a property called \emph{negative association}, a generalization of independence.
The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orthant dependence. %TODO: perhaps citations for these two?
Importantly, standard Chernoff--Hoeffding type bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
Negative association is defined as follows:
\begin{definition}
For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, i.e., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, s.t., $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
\end{definition}

\begin{definition}[Negative Association]\label{def:neg_assoc}
A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
\[
\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
\]
\end{definition}

The following proposition summarizes some of the properties for negatively associated sets of random variables.
\begin{proposition}[\cite{joagdev1983}]\label{pro:neg_dep_props}
Summary of results for negatively associated random variables.
\begin{enumerate}
\item \label{it:neg_dep_props:mult_mono} If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector-valued random variables $X$ and $Y$ are independent, then the union $X_1,\dots,X_n,Y_1,\dots,Y_m$ is a set of negatively associated random variables.
\item If $X=(X_1,\ldots,X_n)$ are negatively associated and $f_1, \dots ,f_m : \mathbb R^n \rightarrow \mathbb R$ are all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
\item If $X_1,\dots,X_n$ are independent then $X_1,\dots,X_n$ are negatively associated.
\item A subset of a negatively associated set of random variables is again negatively associated.
\end{enumerate}
\end{proposition}

These properties illustrate the trade-off between negative association and independence.
For example, Property 3 would be true for independent random variables, even without the condition of monotonicity.
%Of course, on the other hand independence is a stronger property and fewer sets of random variables are independent.
To analyze our new subsampler, the following is an important lemma about negative associated random variables.

\begin{lemma}\label{le:neg_assoc_prod}
Let $X_1,\dots,X_n$ be negatively associated and $f_1,\dots,f_n$ be all non-decreasing (or all non-increasing), non-negative functions, then
\[
  \expect \left[\prod_{i=1}^{n} f_i(X_i)\right] \leq \prod_{i=1}^{n} f_i(\expect [X_i]) \textrm{.}
\]
\end{lemma}
\begin{proof}
This follows from the definition of negative associativity (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
\end{proof}

### 2025-02-25 paper/sec/semra.tex
some space saving
minor edits (.. -> ...)

Importantly, CVM is a \emph{probably-approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
$\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
for parameters $\varepsilon$ and $\delta$,
i.e., the probability that the relative error of $R$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.\footnote{The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques. It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.}

@@

\cref{eq:integral_bind} is crucially used to establish invariants of the form we will use in~\cref{sec:invariants}.
\end{description}
%\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}

>>>>

Importantly, CVM is a \emph{probably-approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
$\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
for parameters $\varepsilon$ and $\delta$,
i.e., the probability that the relative error of $R$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.

\begin{remark}
The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques. It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.
\end{remark}

@@

\cref{eq:integral_bind} is crucially used to establish the invariants we introduce next in~\cref{sec:invariants}.
\end{description}
